---
title: "analysis_quarto"
author: "Arwenn Kummer"
format: html
editor: visual
---

## Data Analysis - user friendly

The data set we will analyse involves the species names and counts of trapped ants, across two different sites - Fynbos and Eucalyptus/Gum trees. The data set contains species counts, species names and quantitative soil surface data, collected using a quadrat placed over the area where each pitfall trap was placed.

## Pre-Analysis work

We will use the packages *tidyverse* and *readxl* to load and analyse our data.

```{r}
install.packages("tidyverse")
install.packages("readxl")

library(tidyverse)
library(readxl)
```

## Data Wrangling, Cleaning and Checks

We have data on soil surface features (quadrat data), species counts and species names. We will read in each data sheet, clean it and reorganize it for analysis.

```{r}
quad.dat <- read.csv("quadrat_data_clean.csv")
view(quad.dat)

pitfall.dat <- read.csv("pitfall_data_clean.csv")
view(pitfall.dat)

```

Exploring the data structure - we want to confirm that variables are the correct format/type (i.e. *chr* for sites, and *int* for the counts).

```{r}
str(quad.dat)
str(pitfall.dat)
```

Next we will confirm that there are no "empty" observations (i.e. missing values), and no duplicates.

```{r}
## check for missing values ----
sum(is.na(quad.dat))
sum(is.na(pitfall.dat))

## check for duplicates ----
sum(duplicated(quad.dat))
sum(duplicated(pitfall.dat))
```

No missing values, and no duplicated values either! Yay! Now we can proceed with Analysis

## Analysis

\
